Practical Machine Learning Project
======
```{r setoptions, echo=FALSE, message=FALSE}
library("knitr")
options(warn=-1)
opts_chunk$set(echo=TRUE, results="asis")
```
##Author: Kevin

##Project task (Question): 
To build a machine learning algorithm to predict activity quality from activity monitors

##Getting input data
1. First we obtain data from this predictive analysis
2. Create local fold "data" to store data and download data files (training set and test set)
3. Load data sets into working space

```{r cache=TRUE}
if (!file.exists("data")) dir.create("data")
setwd("data") ## ==  setwd("./data")

setInternet2(use=T)

fileUrl_trainingData<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
fileUrl_testData<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

download.file(fileUrl_trainingData, destfile="./trainingData.csv")
download.file(fileUrl_testData, destfile="./testData.csv")

list.files("./")
dateDownloaded<-date()

trainingDataSet<-read.csv("trainingData.csv")
testDataSet<-read.csv("testData.csv")
```

##Identify features
any or combination of variables/predictors, decide threshold, etc. 
what is the definition of "good activity quality"?

##Design algorithm
use features and specific logic to make prediction 

##Evaluation: Cross-validation & out-of-sample error estimate


